Thread Pool in GeckoDB/BOLSTER - Implementing a Thread Pool for a High Performance Database System \\

### 2018-07-05 Presentations/Paper/paper_thread_pool.tex
+ some optimizations + evaluation

Each assigned task contains a function pointer to a routine and a pointer to data. We implemented the task queue as a generic priority queue that uses a mutex thread to enqueue tasks threadsafe. Every time a new task is enqueued to the task queue, two scenarios can occur. If the new task has a priority, it is placed in the right slot of the task queue regarding its priority and the subsequent tasks are relocated. If the  new task has no priority, it is appended to the end of the queue. Furthermore, tasks can refer to a handle, which is used by the thread pool to wait for a specific amount of tasks.

\subsection{Processing Tasks}
Processing tasks in a thread pool is a balancing act between generalization the of tasks and the performance of their execution. With higher generalization, the implementation can be used for more use cases, but the execution performance may decrease. Therefore, we provide a set of functions to process task in the thread pool to match different requirements of BOLSTER. In Table 1, we show the functions provided by our thread pool to process tasks. One of the requirements of BOLSTER is waiting until a set of tasks is finished. To achieve at least the same performance as the baseline implementation reaches, we let the main thread first participate in the execution of the tasks and then wait until all tasks with the same handle are finished.

@@

@@

In our setups mostly Sky- and Kabylake processors have been used with memory
ranging from 8 to 24 GB of both DDR3-SODIMM and DDR4-SODIMM. 
All benchmarks have been repeated three times and a median value of all
repetitions is being used for evaluation.

@@

Secondly, the task creation and enqueuing have only slight on the performance, since the baseline does not include these operations, but is still clearly slower than the thread pool implementation.

@@

@@

@@

@@

>>>>

Each assigned task contains a function pointer to a routine and a pointer to data. We implemented the task queue as a generic priority queue utilizing an implicit heap. To ensure thread safety a mutex is used while operating on the queue. Every time a new task is enqueued to the task queue, two scenarios can occur. If the new task has a priority, it is placed in the right slot of the task queue regarding its priority and the subsequent tasks are relocated. If the  new task has no priority, it is appended to the end of the queue. Furthermore, tasks can refer to a handle, which is used by the thread pool to wait for a specific amount of tasks.

With an increasing number of threads interactions with the queue could bottleneck the task throughput. An alternative lock free priority queue \cite{it:2018-003} based on a skip list is considered to improve on this.
\subsection{Processing Tasks}
Processing tasks in a thread pool is a balancing act between generalization the of tasks and the performance of their execution. With higher generalization, the implementation can be used for more use cases, but the execution performance may decrease. Therefore, we provide a set of functions to process task in the thread pool to match different requirements of BOLSTER. In Table 1, we show the functions provided by our thread pool to process tasks. One of the requirements of BOLSTER is waiting until a set of tasks is finished. To achieve at least the same performance as the baseline implementation reaches, we let the main thread first participate in the execution of the tasks and then wait until all tasks with the same handle are finished.

@@

Since $openTaskCount$ has to be altered from different threads the required decrements are ensured to be atomic. This is in opposition to the usual goal of slot-maps to improve cache locality. Atomics on x86 usually prevent their whole cache line from being used while they are changed. Thus a sparse distribution of slots should reduces collisions. To quantify this difference the baseline implementation searches for free slots linearly from the beginning, while an alternative approach tries to distribute the slots over multiple cache lines:
\begin{lstlisting}[style=CStyle]
size_t ind = 0;
// linear search, compact slot usage
for(; ind < pool->task_state_capacity && pool->task_group_states[ind].task_count; ++ind);
// distributed over different cache lines
for(; pool->task_group_states[ind].task_count; ind = (ind + 8) % MAX_NUM_TASKS);
\end{lstlisting}

@@

In our setups Sky- and Kabylake aswell as Ryzen processors have been used with memory
ranging from 8 to 24 GB of both DDR3-SODIMM and DDR4-SODIMM. 
All benchmarks have been repeated three times and a median value of all
repetitions is being used for evaluation.

@@

Secondly, the task creation and enqueuing have only slight impact on the performance, since the baseline does not include these operations, but is still clearly slower than the thread pool implementation.

@@

To further test the performance of the task-completion system, in particular the previously mentioned wait-slot acquisition, the main thread periodicity has to wait for a specific task to finish and the total utilization is measured.

@@

The selection strategy of wait-slots has only a small impact on occupation as seen in \ref{fig8}. The uniform distributions advantage amounts to $0.1\%$ across different numbers of threads and is far overshadowed by differences in the threads to tasks ratio.

Figure \ref{fig9} shows the lock-free priority queue performing better. With a difference in utilization of $2\%$ when using only two threads the improvement over our mutex based implementation is not insignificant. A total occupation of over $99.5\%$ also demonstrates that most overhead in our implementation comes from the priority queue when supplying a balanced stream of tasks. With an increase in available threads the performance of the mutex solution further decreases, while the lock-free queue-variants utilization remains nearly constant.

@@

\begin{figure}
	\includegraphics[width=0.5\textwidth]{img/slot_distr.png}
	\caption{Utilization \% with different slot distribution strategies}
	\label{fig8}
\end{figure}

\begin{figure}
	\includegraphics[width=0.5\textwidth]{img/lock_free.png}
	\caption{Comparison of Utilization \% of the lock-free queue versus mutex based heap}
	\label{fig9}
\end{figure}

@@

The choice of a different priority queue data structure presents an opportunity for further optimizations, in particular to scale well with large amounts of threads.
